---
title: "Getting started with composite selection signals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with composite selection signals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
data.table::setDTthreads(2)
```

```{r setup, message = FALSE}
library(cssig)
library(data.table)
```

## What CSS does

Composite selection signals (CSS) combine several selection tests into one
index. Each constituent test is ranked genome-wide, the ranks become normal
quantiles, and the quantiles are averaged across tests at every SNP. A locus
where several tests agree rises; a locus where only one test is extreme is
pulled back towards the genome-wide average.

The method is due to Randhawa, Khatkar, Thomson and Raadsma
([2014](https://doi.org/10.1186/1471-2156-15-34),
[2015](https://doi.org/10.1534/g3.115.017772)).

## The input

`cssig` starts from **pre-computed constituent statistics**, one row per SNP.
It does not read VCFs or compute XP-EHH; use `rehh`, `selscan` or `hapbin` for
that and bring the results here.

```{r}
data(css_sim_small)
head(css_sim_small[, .(chr, pos, fst, xpehh, ddaf)])
```

These are simulated data with sweeps in known places — see `?css_sim` for how
they were generated, and `css_sim_truth` for the answers.

`css_input()` validates the table and records, per test, which direction means
"more evidence of selection":

```{r}
x <- css_input(css_sim_small,
               tests = c(fst = "high", xpehh = "high", ddaf = "high"))
x
```

## The pipeline

```{r}
res <- css(x)
res <- css_smooth(res, half_width = 5e5, min_snps = 5)
res <- css_threshold(res, top = 0.001)
```

Each stage adds columns to `res`. Because `data.table` assigns by reference, no
copy of the table is made between stages — but the object you passed to
`css_input()` is never touched:

```{r}
"css" %in% names(css_sim_small)   # the original is unchanged
```

If you would rather each stage returned a fresh copy, pass `.copy = TRUE`.

## Calling regions

Two rules are available, one from each paper. `"cluster"` (the 2014 rule) wants
at least three significant SNPs within a window; `"flank"` (the 2015 rule) wants
one top-0.1% SNP flanked by five top-1% neighbours.

```{r}
reg <- css_regions(res, method = "cluster")
summary(reg)
```

The `n_*` columns count how many SNPs in each constituent test's *own* top
fraction fall inside the region. This is the comparison in Table 2 of the 2014
paper, and it is how you tell whether a composite signal is supported by several
tests or driven by one.

## Plots

```{r}
css_manhattan(res, regions = reg)
```

```{r}
css_chrom_plot(res, chr = "2")
```

Points are per-SNP CSS; the line is the sliding-window mean.

```{r fig.height = 6}
if (nrow(reg)) css_region_plot(res, reg[1])
```

## Checking the null

Under no selection the CSS p-values should be uniform. A QQ plot that leaves the
diagonal along its whole length, rather than only at the tail, means the null is
miscalibrated — usually because the constituent tests are strongly correlated.

```{r}
css_qq(res)
```

```{r}
css_test_cor(res)
```

CSS should correlate with each constituent. The constituents need not correlate
much with each other — that they do not is exactly why combining them helps.

## Three things worth knowing

**CSS depends only on ranks.** Standardising XP-EHH or ΔDAF to *N*(0,1), as both
papers do, leaves CSS *exactly* unchanged. It matters only for plotting. More
generally, any strictly increasing transform of a constituent test gives
identical CSS:

```{r}
d2 <- copy(css_sim_small)
d2[, fst := exp(fst * 5)]
r2 <- css(css_input(d2, tests = c(fst = "high", xpehh = "high", ddaf = "high")))
all.equal(res$css, r2$css)
```

**Ranks are genome-wide.** Adding or removing SNPs changes every SNP's CSS. Keep
the SNP set fixed when comparing runs.

**Smoothed scores are not p-values.** `css_smooth()` averages −log10(*p*) over
neighbouring SNPs; the result is a useful score but no longer a p-value, so
`css_fdr()` refuses smoothed input. Estimate FDR first, then smooth:

```{r}
fdr_res <- css_fdr(css(css_input(css_sim_small,
                     tests = c(fst = "high", xpehh = "high", ddaf = "high"))),
                   method = "BH")
sum(fdr_res$qval <= 0.05)
```

**And the raw CSS p-value is not calibrated either.** The formula
*p* = 1 − Φ(√*m* · Z̄) assumes the constituent tests are independent. They are
not, and the correlation is not always positive — in `css_sim`, XP-EHH and ΔDAF
correlate at about −0.32, which *deflates* the variance of Z̄:

```{r}
r_full <- css(css_input(css_sim_small,
                        tests = c(fst = "high", xpehh = "high", ddaf = "high")))
sd(sqrt(3) * r_full$zbar)   # 1.0 only if the tests were independent
```

So treat CSS as a **ranking**, not as a calibrated tail probability. This is why
both papers threshold on the empirical top 0.1% rather than on a p-value, and
why `css_fdr()`'s output deserves a look at `css_qq()` before you lean on it.
See `?css_fdr` for the details.

## Where to go next

- `vignette("simulation")` — how the example data were built, a power comparison
  of CSS against each constituent test on data where the truth is known, and a
  worked demonstration of the `weights` argument.
- `?css_reciprocal` — the signed, two-direction contrast of the 2015 paper.
- `?css_circos` — the circular multi-track plot.

## References

The method is defined in the first of these; the second extends it to complex
traits and adds the reciprocal cohort contrast.

Randhawa IAS, Khatkar MS, Thomson PC, Raadsma HW (2014). Composite selection
signals can localize the trait specific genomic regions in multi-breed
populations of cattle and sheep. *BMC Genetics* **15**:34.
doi:[10.1186/1471-2156-15-34](https://doi.org/10.1186/1471-2156-15-34)

Randhawa IAS, Khatkar MS, Thomson PC, Raadsma HW (2015). Composite selection
signals for complex traits exemplified through bovine stature using multibreed
cohorts of European and African *Bos taurus*. *G3* **5**:1391-1401.
doi:[10.1534/g3.115.017772](https://doi.org/10.1534/g3.115.017772)

```{r}
citation("cssig")
```
