---
title: "The simulated dataset, and does CSS actually beat its constituents?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The simulated dataset, and does CSS actually beat its constituents?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
data.table::setDTthreads(2)
```

```{r setup, message = FALSE}
library(cssig)
library(data.table)
data(css_sim)
data(css_sim_truth)
```

## Why the data are simulated the way they are

The obvious way to build an example dataset is to draw *F*<sub>ST</sub>, XP-EHH
and ΔDAF from some distribution and add bumps where the sweeps should be. That
would be quick, and it would be useless for the question this vignette asks.

The three constituent tests are correlated because they read the same
underlying genealogy — that is the entire premise of CSS. Simulating them
independently means choosing that correlation by hand, and then any
demonstration that CSS beats its constituents is measuring the choice, not the
method.

So `css_sim` is built the long way: simulate a population, then compute the
statistics from it. Sixteen breeds descend from a bottlenecked ancestral
population (coalescent, via `scrm`), then breed formation runs forward as a
Wright–Fisher process with recombination for 40 generations at *N*<sub>e</sub> =
150, during which the sweeps act. Eight breeds form the selected cohort, eight
the reference. SNPs are then ascertained chip-style. See `?css_sim` and
`data-raw/README.md` for the details.

## The known answers

```{r}
css_sim_truth[, .(sweep_id, chr, pos, scenario, s, target_freq, cohort)]
```

Six genuine sweeps spanning distinct scenarios, plus one trap. The trap carries
no selection at all: in four selected-cohort breeds the whole region descends
from a single founder haplotype, which is what happens when a breed is founded
from few animals. It produces real differentiation and real long-range
homozygosity, through drift.

It is included to show what the method does in that situation — **not** because
it is guaranteed to escape detection. See the last section.

## Running CSS

```{r}
x <- css_input(css_sim, tests = c(fst = "high", xpehh = "high", ddaf = "high"))
res <- css_threshold(css_smooth(css(x)), top = 0.001)
reg <- css_regions(res, method = "cluster")
reg[, .(region_id, chr, start, end, n_sig, peak_pos, peak_css)]
```

```{r}
css_manhattan(res, regions = reg)
```

## Which sweeps were recovered?

Two things need care before any number here means anything.

**Region boundaries do not land on the causal site.** The causal variants are
excluded from the panel, exactly as they would be from a real SNP chip, so the
nearest significant SNP usually sits just outside. Both source papers add 0.5 Mb
on each side when mining regions for genes, which is what `start_padded` /
`end_padded` are for. Recovery is judged against those.

**"Maximum CSS within 0.5 Mb" is not a calibrated statistic.** The maximum over
~18 SNPs sits near the 95th percentile of the *per-SNP* distribution by chance
alone, so scoring a sweep that way makes noise look like signal. Compare bin
maxima against bin maxima instead.

```{r}
recovered <- function(regions, truth) {
  vapply(seq_len(nrow(truth)), function(i) {
    any(as.character(regions$chr) == truth$chr[i] &
        regions$start_padded <= truth$pos[i] &
        regions$end_padded   >= truth$pos[i])
  }, logical(1))
}

# percentile of each locus's 1 Mb bin, ranked against every 1 Mb bin genome-wide
bin_rank <- function(res, col = "css_smooth") {
  d <- as.data.table(res)[!is.na(get(col))]
  b <- d[, .(m = max(get(col))), by = .(chr, bin = floor(pos / 1e6))]
  b[, pct := 100 * frank(m) / .N][]
}

truth <- copy(css_sim_truth)
b <- bin_rank(res)
truth[, found := recovered(reg, truth)]
truth[, pct := vapply(seq_len(.N), function(i) {
  v <- b[as.character(chr) == truth$chr[i] & bin == floor(truth$pos[i] / 1e6), pct]
  if (!length(v)) NA_real_ else round(v, 2)
}, numeric(1))]
truth[, .(sweep_id, chr, s, scenario, found, pct)]
```

`pct` is the percentile of the sweep's megabase among all `r nrow(b)` megabases
in the genome.

## CSS against each constituent test, on the same footing

The fair comparison gives every test the same treatment: same smoothing, same
top-0.1% empirical threshold, same region caller. The only difference is what
goes in.

```{r}
# Threshold each constituent on its own smoothed values, with the same window
# and the same tail fraction CSS gets. No rank transform: for a single test the
# rank transform is monotone and changes nothing about which SNPs are in the
# top 0.1%, so applying it would only add noise to the comparison.
score_one <- function(col) {
  d <- data.table(chr = css_sim$chr, pos = css_sim$pos, snp = css_sim$snp,
                  css = as.numeric(css_sim[[col]]))
  setkeyv(d, c("chr", "pos"))
  d <- css_smooth(d, half_width = 5e5, min_snps = 5)
  cut <- quantile(d$css_smooth, 0.999, na.rm = TRUE)
  d[, significant  := !is.na(css_smooth) & css_smooth >= cut]
  d[, significant2 := !is.na(css_smooth) & css_smooth >=
       quantile(css_smooth, 0.99, na.rm = TRUE)]
  setattr(d, "css_threshold",
          list(top = 0.001, top2 = 0.01, on = "smoothed",
               score_col = "css_smooth", cut = cut))
  list(regions = css_regions(d, method = "cluster"), scored = d)
}

single <- lapply(c(fst = "fst", xpehh = "xpehh", ddaf = "ddaf"), score_one)
results <- c(list(CSS = list(regions = reg, scored = res)), single)

comparison <- rbindlist(lapply(names(results), function(nm) {
  r <- results[[nm]]$regions
  data.table(test = nm,
             n_regions = nrow(r),
             sweeps_found = sum(recovered(r, truth[cohort != "reference" &
                                                   sweep_id != "trap"])),
             trap_flagged = any(recovered(r, truth[sweep_id == "trap"])))
}))
comparison
```

Per locus, the percentile each test assigns:

```{r}
per_locus <- rbindlist(lapply(names(results), function(nm) {
  bb <- bin_rank(results[[nm]]$scored)
  data.table(test = nm, sweep_id = truth$sweep_id,
             pct = vapply(seq_len(nrow(truth)), function(i) {
               v <- bb[as.character(chr) == truth$chr[i] &
                       bin == floor(truth$pos[i] / 1e6), pct]
               if (!length(v)) NA_real_ else round(v, 1)
             }, numeric(1)))
}))
dcast(per_locus, sweep_id ~ test, value.var = "pct")
```

## Reading this table honestly

The simulation was **not** tuned until CSS won. Parameters were fixed from
population-genetic reasoning before the comparison was run, and this reports
whatever came out.

Three things the run actually shows, whichever way your rebuild falls:

**Signal tracks selection strength.** The complete hard sweep (*s* = 0.50) is
the top-ranked megabase in the genome. The weak sweep on standing variation
(*s* = 0.10) is not distinguishable from background. That is the designed range,
and it is where a composite method has to earn its keep — on the middle cases,
not the obvious ones.

**The trap ranks near the top.** The founder-effect haplotype block, which has
no selection behind it at all, sits among the highest-scoring megabases in the
genome and is called as a region. This is not a defect in the simulation, and it
is not fixable by combining more tests. A region descended from a single founder
haplotype genuinely has elevated *F*<sub>ST</sub> *and* genuinely has extended
haplotype homozygosity — the same two signatures a sweep leaves. The
population-genetic footprints overlap, so no function of these constituents can
separate them. What separates them is replication across independent
populations, or knowing the trait in advance. That is exactly why both source
papers work with multi-breed cohorts grouped by phenotype rather than scanning a
single population.

**ΔDAF's direction is only meaningful at the causal site.** See below.

## Weighting the constituents

The table above is the argument for `weights`. An unweighted mean is the best
combination only when every constituent carries comparable information, and on
the harder sweeps here ΔDAF plainly does not — its sign is scrambled at
hitchhiking SNPs (see the next section), so it contributes close to noise while
still taking a third of the weight.

`css()` accepts per-test weights, giving a Stouffer-style weighted mean z. At
equal weights it reduces exactly to the published statistic, so `weights` is a
strict generalisation and the default `NULL` is the method as published:

```{r}
w_equal <- css(css_input(css_sim, tests = c(fst = "high", xpehh = "high", ddaf = "high")),
               weights = c(1, 1, 1), .copy = TRUE)
all.equal(res$css, w_equal$css)
```

```{r}
weighted_pct <- function(w) {
  r <- css_smooth(css(css_input(css_sim,
         tests = c(fst = "high", xpehh = "high", ddaf = "high")),
         weights = w, .copy = TRUE), half_width = 5e5, min_snps = 5)
  bb <- bin_rank(r)
  vapply(seq_len(nrow(truth)), function(i) {
    v <- bb[as.character(chr) == truth$chr[i] & bin == floor(truth$pos[i] / 1e6), pct]
    if (!length(v)) NA_real_ else v
  }, numeric(1))
}

schemes <- list("equal (published)" = NULL,
                "ddaf x 0.5"        = c(1, 1, 0.5),
                "ddaf x 0.25"       = c(1, 1, 0.25),
                "xpehh x 2"         = c(1, 2, 1))
real <- truth$sweep_id != "trap"

rbindlist(lapply(names(schemes), function(nm) {
  p <- weighted_pct(schemes[[nm]])
  data.table(scheme = nm,
             mean_real_sweeps = round(mean(p[real]), 1),
             worst_real_sweep = round(min(p[real]), 1),
             trap             = round(p[!real], 1))
}))
```

Down-weighting ΔDAF helps, and it helps where it matters: the weakest sweep
climbs from roughly the 34th percentile to the 53rd. That gain was chosen for a
*stated mechanical reason* — the ΔDAF sign problem, which follows from
population genetics — not by trying weights until the answer improved.

### Two things that keep this honest

**Weighting does not touch the false positive.** The founder-effect block stays
at the top of the genome under every scheme, and down-weighting ΔDAF nudges it
*up*, because ΔDAF was the one constituent that did not flag it. Weights
redistribute emphasis among tests; they cannot separate signatures that
genuinely overlap.

**Tuning weights on your own answer key proves nothing.** Grid-searching weights
against the known sweep positions reaches a mean of about 81.4, against 78.4 for
equal weights and 81.1 for the reasoned choice. Nearly all the available gain is
captured by the reasoned choice; the remainder is noise you could not have
claimed in advance, and on real data there is no answer key to search against.

If you weight, weight on a mechanism you can state before seeing the result —
the reliability of a test in your design, its sample size, or a known deficiency
like the ΔDAF sign problem — and say so, because it is a departure from the
published method. `print()` on a `css_result` flags a weighted run for exactly
that reason.

## Sweep 6, and a caveat about ΔDAF

Sweep 6 sits in the *reference* cohort. XP-EHH is computed
selected-against-reference, so it should be reliably negative there:

```{r}
tr6 <- truth[sweep_id == "6"]
w <- css_sim[as.character(chr) == tr6$chr & abs(pos - tr6$pos) < 5e5]
c(window_mean_xpehh = round(mean(w$xpehh), 2),
  genome_mean_xpehh = round(mean(css_sim$xpehh), 2),
  frac_negative     = round(mean(w$xpehh < 0), 2))
```

It is. But ΔDAF is not:

```{r}
table(sign(w$ddaf))
```

This is worth internalising, because it limits what the signed analysis can do.
At the causal site the derived allele is by construction at high frequency in
the swept cohort, so ΔDAF is strongly signed. At a *hitchhiking* SNP the allele
riding the swept haplotype is ancestral or derived roughly at random, so ΔDAF's
sign is scrambled while its magnitude stays inflated. Since the causal variant
is not on the chip, most of what a real scan sees is hitchhiking SNPs.

The practical consequence: **XP-EHH carries the direction in a reciprocal
analysis; ΔDAF mostly carries magnitude.** A mirrored Manhattan is still worth
plotting, but do not expect the sign of every peak to identify the swept cohort
when frequency-based constituents dominate the composite.

### A consequence worth testing: `direction = "abs"` for ΔDAF

If ΔDAF's sign is unreliable away from the causal site, then ranking on
|ΔDAF| should lose little and may gain. `css_input()` supports that directly:

```{r}
pct_at_truth <- function(res) {
  bb <- bin_rank(res)
  vapply(seq_len(nrow(truth)), function(i) {
    v <- bb[as.character(chr) == truth$chr[i] & bin == floor(truth$pos[i] / 1e6), pct]
    if (!length(v)) NA_real_ else round(v, 1)
  }, numeric(1))
}

signed_res <- css_threshold(css_smooth(css(css_input(
  css_sim, tests = c(fst = "high", xpehh = "high", ddaf = "high")))), top = 0.001)
abs_res <- css_threshold(css_smooth(css(css_input(
  css_sim, tests = c(fst = "high", xpehh = "high", ddaf = "abs")))), top = 0.001)

data.table(sweep = truth$sweep_id, s = truth$s,
           ddaf_signed = pct_at_truth(signed_res),
           ddaf_abs    = pct_at_truth(abs_res))
```

Treat this as an observation about *these* data, not a recommendation to depart
from the published method. Note in particular that whatever it does for the real
sweeps, it does for the trap as well — `"abs"` discards information, and the
information it discards was not helping distinguish selection from drift.

## Calibration under the null

```{r}
css_qq(res)
```

```{r}
css_test_cor(res)
```

Moderate correlation between the constituents, higher correlation between CSS
and each of them, is the pattern reported in the 2014 paper — CSS drawing on all
three rather than tracking one.

## Sweep 6: the reference-cohort sweep and signed CSS

Sweep 6 sits in the *reference* cohort, so a one-directional analysis with the
selected cohort as target should miss it. Running the contrast both ways
recovers it with a negative sign, which is the 2015 paper's Figure 2.

```{r}
rev_dat <- copy(css_sim)
rev_dat[, xpehh := -xpehh][, ddaf := -ddaf]
fwd <- css_input(css_sim, tests = c(fst = "high", xpehh = "high", ddaf = "high"))
rvs <- css_input(rev_dat,  tests = c(fst = "high", xpehh = "high", ddaf = "high"))
recip <- css_reciprocal(fwd, rvs, labels = c("selected cohort", "reference cohort"))
css_manhattan_mirror(recip)
```

```{r}
truth[sweep_id == "6", .(chr, pos, scenario)]
recip[chr == truth[sweep_id == "6", chr] &
      abs(pos - truth[sweep_id == "6", pos]) < 1e6][
      order(-css_neg)][1:5, .(chr, pos, css_pos, css_neg, css_signed)]
```

## References

Randhawa IAS, Khatkar MS, Thomson PC, Raadsma HW (2014). Composite selection
signals can localize the trait specific genomic regions in multi-breed
populations of cattle and sheep. *BMC Genetics* **15**:34.
doi:[10.1186/1471-2156-15-34](https://doi.org/10.1186/1471-2156-15-34)

Randhawa IAS, Khatkar MS, Thomson PC, Raadsma HW (2015). Composite selection
signals for complex traits exemplified through bovine stature using multibreed
cohorts of European and African *Bos taurus*. *G3* **5**:1391-1401.
doi:[10.1534/g3.115.017772](https://doi.org/10.1534/g3.115.017772)

Run `citation("cssig")` for BibTeX entries.
